---
title: "Reproducible Research: Peer Assessment 1"
author: mb84
---
## Introduction 

The variables included in this dataset are:

- steps:      Number of steps taking in a 5-minute interval (missing values are coded as NA)
- date:       The date on which the measurement was taken in YYYY-MM-DD format
- interval:   Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

## Loading and preprocessing the data


```{r, importdata, echo=TRUE, warning=FALSE}

library(data.table)
unzip(zipfile = "activity.zip")
adata <- read.csv(file = "activity.csv", header=TRUE)
adatadt <- data.table(adata)

```

## What is mean total number of steps taken per day?

For this part of the assignment, you can ignore the missing values in the dataset.  

1. **Calculate the total number of steps taken per day**

```{r, question1.1, cache=FALSE, echo=TRUE}

total_xday <- adatadt[,sum(steps), by = date]
setnames(total_xday,"V1","TotalSteps")
head(total_xday)

```


2. **If you do not understand the difference between a histogram and a barplot, research the difference between them. Make a histogram of the total number of steps taken each day**  

```{r, question1.2, cache=FALSE, echo=TRUE}

hist(total_xday$TotalSteps, main = "Total number of steps taken each day", xlab = "Total steps")

```

3. **Calculate and report the mean and median of the total number of steps taken per day**

```{r, question1.3, cache=FALSE, echo=TRUE}

mean_totxday <- mean(total_xday$TotalSteps, na.rm = T)
median_totxday <- median(total_xday$TotalSteps, na.rm = T)

```


The mean total number of steps taken per day and rounded to integer is `r as.integer(mean_totxday)` whereas the median is `r as.integer(median_totxday)`.

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
